---
title: "Alteromonas_3m2ob_uptake"
author: "KHH"
date: "12/13/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Goals

Analyze the growth of Alteromonas macleodii on 3-methyl-2-oxobutanoic acid (3m2ob), and the change in 3m2ob concentration during Alteromonas growth. 

Though initial curve fitting and quantification was performed in MATLAB (see github.com/KujawinskiLaboratory for "considerMAVEN.m" and "riMAVEN.m" scripts), additional data visualization and QA/QC was performed here. Additionally, quantification using the ratio of unlabeled to labeled 3m2ob was performed here (considerMAVEN.m and riMAVEN.m use internal standards for QA/QC, but not for normalization and quantification)
 
# Get packages and functions, and set paths

```{r get packages}
#for data management and plotting
require(tidyverse)

#some useful color palettes 
require(ggsci)
require(pals)
require(PaletteWoodsHole)
```


```{r get summary function}
## Summarizes data.
## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}

#example use:
#d_summary_name <- summarySE(d_example, measurevar = "variable to summarize", groupvars = c("group by this", "group by that"))
```


```{r define a function to save plots}
save_plots <- function(p, filename, path){
  ggsave(filename = filename, path = path, plot = p, device = "pdf", dpi = "retina", bg = "transparent")
}
```


```{r}
##Set your path for saving plots
path_for_saving_plots <- "~/Documents/MIT_WHOI_JP/Kuj:Chisholm data/Uptake experiments/04.2024 reanalysis/3m2ob/Figures/"

```

# Get data

1. A spreadsheet of peak areas in all samples and standards for the compounds targeted for LCMS (3-methyl-2-oxobutanoic acid, D5-3-methyl-2-oxobutanoic acid, and D2-biotin). 
2. A spreadsheet of sample volumes and cell concentrations. 


```{r pull data into R}
Uptake_data <- "~/Documents/MIT_WHOI_JP/Kuj:Chisholm data/Uptake experiments/3m2ob uptake 7 reanalysis/KHU7_peaklist_v3_peakAreaIS_workup_2023.csv"

d_3m2ob_conc <- read.table(Uptake_data, header = TRUE, sep = ",", stringsAsFactors = FALSE)
```


```{r snag dataframe wtih sample volume info}
volume_data <- "~/Documents/MIT_WHOI_JP/Kuj:Chisholm data/Uptake experiments/KHU7/KHU7_forR.csv"

d_3m2ob_volume_full <- read.table(volume_data, header = TRUE, sep = ",", stringsAsFactors = FALSE)
```


```{r add detail to Sample.condition}
# 
# if(d_3m2ob_conc$Sample.name == "KHU7_Pool"){
#   d_3m2ob_conc$Sample.condition <- "pool"
# }


d_3m2ob_conc[(d_3m2ob_conc$Sample.name == "KHU7_Pool"), "Sample.condition"] <- "pool"
d_3m2ob_conc[(d_3m2ob_conc$Sample.name == "PoolCondition"), "Sample.condition"] <- "pool"
d_3m2ob_conc[(d_3m2ob_conc$Sample.name == "MQ Blank" | d_3m2ob_conc$Sample.name == "MQ blank pos"), "Sample.condition"] <- "MQ blank"
d_3m2ob_conc[(d_3m2ob_conc$injection.number > 4 & d_3m2ob_conc$injection.number < 16), "Sample.condition"] <- "MQ curve"
d_3m2ob_conc[(d_3m2ob_conc$injection.number > 108 & d_3m2ob_conc$injection.number < 121), "Sample.condition"] <- "Matrix curve"
d_3m2ob_conc[(d_3m2ob_conc$injection.number == 2), "Sample.condition"] <- "MQ blank"
d_3m2ob_conc[(d_3m2ob_conc$injection.number == 3), "Sample.condition"] <- "MQ 1000 ng_mL"

#d_OD600_diauxic_analysis_viz[(d_OD600_diauxic_analysis_viz$Carbon_source == "None"), "metabolite"] <- NA
```


```{r plot 3m2ob peak area by injection number}
p <- ggplot(data = d_3m2ob_conc, aes(y = log(X3m2ob.peakArea), x = injection.number, color = Sample.condition))

p + geom_point()
```

```{r}
p <- ggplot(data = d_3m2ob_conc, aes(y = X3m2ob.peakArea, x = injection.number, color = Sample.condition))

p + geom_point()
```


```{r calculate peak area ratios and plot by injection number}
d_3m2ob_conc$peak_area_ratio <- d_3m2ob_conc$X3m2ob.peakArea/d_3m2ob_conc$X3m2ob.d5.peakArea

p <- ggplot(data = d_3m2ob_conc, aes(y = peak_area_ratio, x = injection.number, color = Sample.condition))
p + geom_point()
```


```{r}
p <- ggplot(data = d_3m2ob_conc, aes(y = log(peak_area_ratio), x = injection.number, color = Sample.condition))
p + geom_point()
```

From these plots, it seems like I can leave out the 1000 ng/mL point in my standard curve without affecting quantification too much. I'll try it. 

#Check internal standards - do they look like I expect them to, more or less?

```{r plot 3m2ob-d5 by injection}
p <- ggplot(data = d_3m2ob_conc, aes(y = X3m2ob.d5.peakArea, x = injection.number))

p + geom_point()
```

```{r identify which samples have the anomolous IS signal}

d_3m2ob_IS_outliers <- filter(d_3m2ob_conc, X3m2ob.d5.peakArea < 5000 | X3m2ob.d5.peakArea > 15000)
d_3m2ob_IS_outliers <- d_3m2ob_IS_outliers %>% select(sample, injection.number, Sample.name, X3m2ob.peakArea, X3m2ob.d5.peakArea)
```

My MQ blanks have a very low IS signal, which is as expected, since I didn't put an internal standard in there. 

My 1000 ng/mL samples are the samples with the very high IS signal. Not the dream! I'll see how it looks when I include those samples in a standard curve. 

Out of curiosity, how does the biotin signal look? 

```{r plot d2 biotin by injection}
p <- ggplot(data = d_3m2ob_conc, aes(y = d2.Biotin.peakArea, x = injection.number))

p + geom_point()
```


# Have a look at your standard curves

```{r standard curve in MQ, from peak area}

d_curve_MQ <- filter(d_3m2ob_conc, Std.conc._ng_mL_ >= 0 & injection.number < 20)
d_curve_MQ <- d_curve_MQ %>% select(sample, injection.number, Sample.name, Std.conc._ng_mL_, X3m2ob.peakArea, X3m2ob.d5.peakArea, d2.Biotin.peakArea)

p <- ggplot(d_curve_MQ, aes(y = X3m2ob.peakArea, x = Std.conc._ng_mL_))

p + geom_point()
```

```{r calculate peak area ratios, plot standard curve}
d_curve_MQ$peak_area_ratio <- d_curve_MQ$X3m2ob.peakArea/d_curve_MQ$X3m2ob.d5.peakArea

p <- ggplot(data = d_curve_MQ, aes(y = peak_area_ratio, x = Std.conc._ng_mL_))
p + geom_point()
```

Yeah definitely some weirdos in the standard curve when I plot it this way! Looks like the 1000 ng/mL sample must have gotten a double dose of the internal standard. 

What are the other weirdos?

```{r}
d_curve_MQ %>% filter(peak_area_ratio > 2) %>% list()
```

Ah those are just my zeros! Which have neither metabolite nor IS in there, so it makes sense that they're weirdly high. 

#Fit curve - get R2 and coefficients

```{r look at MQ curve}
#Don't include 0 or 1000 ng/mL endpoints, since those are wonky

d_curve_MQ <- filter(d_curve_MQ, Std.conc._ng_mL_ > 0 & Std.conc._ng_mL_ < 1000)

p <- ggplot(data = d_curve_MQ, aes(y = peak_area_ratio, x = Std.conc._ng_mL_))
p + geom_point() + scale_y_log10() + scale_x_log10()
```

Curious about whether using this standard curve improves quantification substantially relative to using a curve without the internal standard. So for now I'm just going to construct a curve using points down to 10 ng/mL 

```{r narrow down points to use in MQ curve}
d_curve_MQ <- filter(d_curve_MQ, Std.conc._ng_mL_ >= 10 & Std.conc._ng_mL_ < 1000)

p <- ggplot(data = d_curve_MQ, aes(y = peak_area_ratio, x = Std.conc._ng_mL_))
p + geom_point() + scale_y_log10() + scale_x_log10()
```

```{r fit MQ curve}
fit_MQ <- lm(peak_area_ratio ~ Std.conc._ng_mL_, data = d_curve_MQ)
#summary(fit_MQ)
d_MQ_fit_coef <- as.data.frame(t(fit_MQ$coefficients))
```

```{r turn MQ fit coefficients into useful format}
d_MQ_fit_coef$Matrix_type <- "MQ"
d_MQ_fit_coef <- d_MQ_fit_coef %>% dplyr::rename("fit_from_curve_b" = "(Intercept)")
d_MQ_fit_coef <- d_MQ_fit_coef %>% dplyr::rename("fit_from_curve_m" = "Std.conc._ng_mL_")
colnames(d_MQ_fit_coef)
```

```{r look at MM curve}
d_curve_MM <- d_3m2ob_conc %>% filter(injection.number > 108 & injection.number < 121)

p <- ggplot(data = d_curve_MM, aes(y = peak_area_ratio, x = Std.conc._ng_mL_))
p + 
  geom_point() + 
  scale_x_log10() + 
  scale_y_log10()
```
From this plot, it looks like for the matrix matched curve I should only consider standards > 1 ng/mL and < 1000 ng/mL. 

```{r narrow down points for MM curve and plot again}
d_curve_MM <- d_curve_MM %>% filter(Std.conc._ng_mL_ > 1 & Std.conc._ng_mL_ < 1000)

p <- ggplot(data = d_curve_MM, aes(y = peak_area_ratio, x = Std.conc._ng_mL_))
p + 
  geom_point() + 
  scale_x_log10() + 
  scale_y_log10()
```

```{r fit MM curve}
fit_MM <- lm(peak_area_ratio ~ Std.conc._ng_mL_, data = d_curve_MM)
summary(fit_MM)
d_MM_fit_coef <- as.data.frame(t(fit_MM$coefficients))
```

```{r turn MM fit coefficients into useful format}
d_MM_fit_coef$Matrix_type <- "MM"
d_MM_fit_coef <- d_MM_fit_coef %>% dplyr::rename("fit_from_curve_b" = "(Intercept)")
d_MM_fit_coef <- d_MM_fit_coef %>% dplyr::rename("fit_from_curve_m" = "Std.conc._ng_mL_")
colnames(d_MM_fit_coef)
```

```{r merge coefficient calculation dataframes}
d_curve_fits <- rbind(d_MM_fit_coef, d_MQ_fit_coef)
```

#Calculate concentrations in samples from curve

```{r find peak ratio to conc coefficients}
d_curve_fits$peak_ratio_to_conc_m <- 1/d_curve_fits$fit_from_curve_m
d_curve_fits$peak_ratio_to_conc_b <- -1 * (d_curve_fits$fit_from_curve_b/d_curve_fits$fit_from_curve_m)
d_coef_MM <- d_curve_fits %>% filter(Matrix_type == "MM")
d_coef_MQ <- d_curve_fits %>% filter(Matrix_type == "MQ")
```

```{r find concentrations from peak ratio coefficients}
MM_m <- d_coef_MM$peak_ratio_to_conc_m
MM_b <- d_coef_MM$peak_ratio_to_conc_b
MQ_m <- d_coef_MQ$peak_ratio_to_conc_m
MQ_b <- d_coef_MQ$peak_ratio_to_conc_b

d_3m2ob_conc$MM_m <- MM_m
d_3m2ob_conc$MM_b <- MM_b
d_3m2ob_conc$MQ_m <- MQ_m
d_3m2ob_conc$MQ_b <- MQ_b

d_3m2ob_conc$instrument_conc_MM <- d_3m2ob_conc$peak_area_ratio * d_3m2ob_conc$MM_m + d_3m2ob_conc$MM_b

d_3m2ob_conc$instrument_conc_MQ <- d_3m2ob_conc$peak_area_ratio * d_3m2ob_conc$MQ_m + d_3m2ob_conc$MQ_b
```


#Convert from instrument conc to sample conc

```{r}
d_3m2ob_volume_full <- d_3m2ob_volume_full %>% dplyr::rename("Sample.name" = "Sample_name")

d_3m2ob_volume <- select(d_3m2ob_volume_full, c("Sample.name", "Cells_mL", "Volume_filtrate_mL"))
```

```{r merge sample volume info w/ instrumental conc dataframe}
d_3m2ob_conc <- full_join(d_3m2ob_conc, d_3m2ob_volume, by = "Sample.name")
```

In this experiment, extracted samples were dried down, brought up in 1 mL of MQ, and diluted 400X prior to analysis. The extraction efficiency of 3m2ob is 10.1%. 100/10.1 = 9.90099; 400*9.0099 = 3960.396. Multiply calculated instrumental concs by 3960.396 to get to pre-extraction ng 3m2ob. 

Then, divide by mL filtrate to get to sample ng/mL 3m2ob. 

Then, the MW of 3m2ob-Na (the compound used for making up 3m2ob std curve) is 138.10 g/mol. Divide by 138.10 to get to nmol/mL = µmol/L = µM

```{r convert from instrument conc to sample conc}
d_3m2ob_conc$MM_pre_extraction_ng <- d_3m2ob_conc$instrument_conc_MM * 3960.396
d_3m2ob_conc$MM_sample_conc_ng_mL <- d_3m2ob_conc$MM_pre_extraction_ng / d_3m2ob_conc$Volume_filtrate_mL
d_3m2ob_conc$MM_sample_conc_uM <- d_3m2ob_conc$MM_sample_conc_ng_mL / 138.10

d_3m2ob_conc$MQ_pre_extraction_ng <- d_3m2ob_conc$instrument_conc_MQ * 3960.396
d_3m2ob_conc$MQ_sample_conc_ng_mL <- d_3m2ob_conc$MQ_pre_extraction_ng / d_3m2ob_conc$Volume_filtrate_mL
d_3m2ob_conc$MQ_sample_conc_uM <- d_3m2ob_conc$MQ_sample_conc_ng_mL / 138.10
```


#Plot peak ratio-generated values vs. time

```{r calculate summary values from calculated uM concs}
d_summary_uM_MM <- summarySE(d_3m2ob_conc, measurevar = "MM_sample_conc_uM", groupvars = c("t", "Sample.condition"))

d_summary_uM_MQ <- summarySE(d_3m2ob_conc, measurevar = "MQ_sample_conc_uM", groupvars = c("t", "Sample.condition"))
```


```{r plot concentration over time using MM data}
p <- ggplot(data = filter(d_summary_uM_MM, Sample.condition == "Treatment"), aes(y = MM_sample_conc_uM, x = t, ymin = MM_sample_conc_uM - sd, ymax = MM_sample_conc_uM + sd))

p + geom_point() + geom_errorbar()
```


```{r plot concentration over time using MQ data}
p <- ggplot(data = filter(d_summary_uM_MQ, Sample.condition == "Treatment"), aes(y = MQ_sample_conc_uM, x = t, ymin = MQ_sample_conc_uM - sd, ymax = MQ_sample_conc_uM + sd))

p + geom_point() + geom_errorbar()
```

#Compare quantification with vs. without an internal standard

```{r generate summary data}
#these are the concentration values calculated with Kujawinski lab code in MATLAB, using riMAVEN and considerMAVEN. These calculations did not use an internal standard for normalization and quantification

d_summary_uM_old <- summarySE(d_3m2ob_volume_full, measurevar = "uM_3m2ob", groupvars = c("t", "Sample_condition"))
```

```{r plot data based on old workflow}
p <- ggplot(data = filter(d_summary_uM_old, Sample_condition == "Treatment"), aes(y = uM_3m2ob, x = t, ymin = uM_3m2ob - sd, ymax = uM_3m2ob + sd))

p + geom_point() + geom_errorbar()
```

In general, using the IS-normalized data results in less variability than using unnnormalized data, as expected. However, at t = 36, there's much less variability in my old calculated concentrations than in my new calcualted concentrations and I am not sure why. Regardless, I think I need to move on - I'll use the IS-normalized, MM concentration data here. 

#Plot IS-normalized, MM conc x time
I'm not sure log scale is very useful for looking at the change in metabolite concentration, so I'll save a version in log and in linear scale. 

```{r}
p_mtab_summary <- ggplot(data = filter(d_summary_uM_MM, Sample.condition == "Treatment"), 
            aes(x = t, y = MM_sample_conc_uM, ymin = MM_sample_conc_uM - sd, ymax = MM_sample_conc_uM + sd)) 

p <- p_mtab_summary + 
  geom_point(size = 4, color = "#1f78b4", shape = 15) + 
  geom_errorbar(color = "#1f78b4") + 
  theme_classic() + 
  theme(text = element_text(size =22)) + 
  xlab("Time (hrs)") + 
  ylab("[3m2ob] (µM)") + 
  ylim(0, 200)

save_plots(p, "uM_3m2ob_KHU7_linearScale.pdf", path_for_saving_plots)

p <- p_mtab_summary + 
  geom_point(size = 4, color = "#1f78b4", shape = 15) + 
  geom_errorbar(color = "#1f78b4") + 
  theme_classic() + 
  theme(text = element_text(size =22)) + 
  xlab("Time (hrs)") + 
  ylab("[3m2ob] (µM)") + 
  scale_y_log10()

save_plots(p, "uM_3m2ob_KHU7_logScale.pdf", path_for_saving_plots)
```

#Plot cells/mL x t

```{r summarize cell conc data}
d_summary_cells_mL <- summarySE(d_3m2ob_conc, "Cells_mL", groupvars = c("t", "Sample.condition"))

d_summary_cells_toPlot <- d_summary_cells_mL %>% 
  filter(Sample.condition == "Treatment")
```

```{r}
p_cells_summary <- ggplot(data = d_summary_cells_toPlot, aes(x = t, y = Cells_mL, ymin = Cells_mL - sd, ymax = Cells_mL + sd))

#plot as you have before, but add in fit curve
p <- p_cells_summary + geom_point(size = 4, color = "#33a02c") +
  geom_errorbar(color = "#33a02c") + 
  geom_smooth(method = "lm", se = FALSE, color = "#33a02c") +
  theme_classic() + 
  theme(text = element_text(size =22)) + 
  xlab("Time (hrs)") + 
  ylab("Cells/mL") + 
  #ylim(0, 5e+07) + 
  scale_y_log10(limits = c(1e+05, 5e+07))

save_plots(p, "cells_mL_3m2ob_KHU7_logScale.pdf", path_for_saving_plots)
```

#Find parameters for fit of cell count ~ time

```{r}
m_cells <- lm(log10(Cells_mL) ~ t, data = d_summary_cells_toPlot)
summary(m_cells)
```



#Calculate the fold change in both 3m2ob and in cells

```{r change in mtab conc}
d_uM_treatment <- d_summary_uM_MM %>%
  filter(Sample.condition == "Treatment")

d_uM_initial <- d_uM_treatment %>%
  filter(t == 0)

d_uM_final <- d_uM_treatment %>%
  filter(t == 58.75)

uM_initial <- d_uM_initial$MM_sample_conc_uM
uM_final <- d_uM_final$MM_sample_conc_uM

fold_decrease_uM <- uM_initial/uM_final

print(fold_decrease_uM)
```

3m2ob saw an 8-fold decrease in concentration

```{r change in cell conc}
d_cell_initial <- d_summary_cells_toPlot %>%
  filter(t == 0)

d_cell_final <- d_summary_cells_toPlot %>%
  filter(t == 58.75)

cells_initial <- d_cell_initial$Cells_mL
cells_final <- d_cell_final$Cells_mL

fold_increase_cells <- cells_final/cells_initial

print(fold_increase_cells)
```

Cells saw a 94-fold increase in concentration

#Plot the no mtab cell growth

```{r}
d_no_mtab <- d_summary_cells_mL %>%
  filter(Sample.condition == "No 3m2ob")
```


```{r}
p_cells_summary_no_mtab <- ggplot(data = d_no_mtab, 
            aes(x = t, y = Cells_mL, ymin = Cells_mL-sd, ymax = Cells_mL+sd)) 

require(scales)

p <- p_cells_summary_no_mtab + 
  geom_point(size = 4, color = "#33a02c") + 
  geom_errorbar(color = "#33a02c") + 
  theme_classic() + 
  theme(text = element_text(size =22)) + 
  xlab("Time (hrs)") + 
  ylab("Cells/mL") + 
  scale_y_log10(labels = scientific, limits = c(1e+05, 5e+07))

p

save_plots(p, "cells_mL_no_mtab_3m2ob_logScale.pdf", path_for_saving_plots)
```

