---
title: "3m2op_Alteromonas_growth"
author: "KHH"
date: "4/10/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals

Analyze growth of Alteromonas macleodii on 3-methyl-2-oxopentanoic acid ± pyruvate. 

In this analysis we tested different ways of dealing with edge effects, which impacted the growth of Alteromonas grown on the outside edges of 96-well plates. We found that simply removing wells from the two rows/columns around the outside of the plate resulted in an insufficient sample size. Instead, we blank corrected, plotted wells individually within a culture condition, and removed edge rows/columns with wells which showed evidence of edge effects. Edge effects typically appeared as cultures which "crashed" immediately after reaching their max OD600, rather than entering stationary phase. 

We next blank corrected, plotted data, found the carrying capacity for each culture, and statistically compared carrying capacities between culture conditions. 

#Get likely required packages and functions


```{r get packages}
#for data management and plotting
require(tidyverse)

#some useful color palettes 
require(ggsci)
require(pals)
require(PaletteWoodsHole)
```


```{r get summary function}
## Summarizes data.
## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}

#example use:
#d_summary_name <- summarySE(d_example, measurevar = "variable to summarize", groupvars = c("group by this", "group by that"))
```


```{r}
##Set your path for saving plots
path_for_saving_plots <- "~/Documents/MIT_WHOI_JP/Kuj:Chisholm data/Growth curves/04.2024_reanalysis/3m2op/Figures/"

  
##Set your metabolite name
mtab_full_name <- "3m2op"
mtab_short_name <- "3m2op"

```


```{r define a function to save plots}
save_plots <- function(p, filename, path){
  ggsave(filename = filename, path = path, plot = p, device = "pdf", dpi = "retina", bg = "transparent")
}
```


## Get data

This analysis requires two dataframes: the OD600 data pulled from the plate reader, and a spreadsheet that describes culture condition by well position. 

```{r get data into R and rearrange}
#Edit the paths below to get to 1. raw OD600 data and 2. a data table that contains condition descriptions. That condition description file should include the following column headers: well, Condition.number, Condition, Condition.name, Total_uM_C, Total_mM_C, Carbon_source, Diauxic. 

OD600_data <- "~/Documents/MIT_WHOI_JP/Kuj:Chisholm data/Growth curves/Metabolite addition/3m2op/3m2op_2/KHH01_089_3m2op_2_forR.csv"

culture_condition_data <- "~/Documents/MIT_WHOI_JP/Kuj:Chisholm data/Growth curves/Metabolite addition/3m2op/3m2op_2/3m2op_2_conditions_description.csv"

#Now get those data into R and into useful formats

d_OD600 <- read.table(OD600_data, header = TRUE, sep = ",", stringsAsFactors = FALSE)
d_culture_conditions <- read.table(culture_condition_data, header = TRUE, sep = ",", stringsAsFactors = FALSE)

d_OD600_long <- gather(d_OD600, key = "well", value = "OD600", "A1", "A2", "A3", "A4", "A5", "A6", "A7", "A8", "A9", "A10", "A11", "A12", "B1", "B2", "B3", "B4", "B5", "B6", "B7", "B8", "B9", "B10", "B11", "B12", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "C9", "C10", "C11", "C12", "D1", "D2", "D3", "D4", "D5", "D6", "D7", "D8", "D9", "D10", "D11", "D12", "E1", "E2", "E3", "E4", "E5", "E6", "E7", "E8", "E9", "E10", "E11", "E12", "F1", "F2", "F3", "F4", "F5", "F6", "F7", "F8", "F9", "F10", "F11", "F12", "G1", "G2", "G3", "G4", "G5", "G6", "G7", "G8", "G9", "G10", "G11", "G12", "H1", "H2", "H3", "H4", "H5", "H6", "H7", "H8", "H9", "H10", "H11", "H12")

d_gc <- full_join(d_OD600_long, d_culture_conditions, by = "well")
```

Our experimental design included a lower concentration metabolite addition, but we decided it did not contribute to our understanding of Alteromonas metabolite uptake, and so that condition was not considered in the manuscript. 

```{r filter out the lowest C metabolite condition}
#This also removes border wells, etc., that have "NA" as their condition
d_gc <- filter(d_gc, Total_mM_C != 0.45)
```

#Do a blank correction

```{r make dataframe for data from blanks}
d_blank_OD600 <- filter(d_gc, Condition == "A")
```

```{r Calculate average blank OD600 at each timepoint}
## Summarizes data.
## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}

d_summary_blank_OD600 <- summarySE(d_blank_OD600, measurevar = "OD600", groupvars = c("time"))
```

```{r add average blank values back into main dataframe and calc corrected OD600}
d_summary_blank_OD600 <- dplyr::rename(d_summary_blank_OD600, "blank_OD600" = "OD600")
d_summary_blank_OD600 <- dplyr::select(d_summary_blank_OD600, "time", "blank_OD600")
d_gc_blank_corr <- full_join(d_gc, d_summary_blank_OD600, by = "time")
d_gc_blank_corr$OD600_corr <- d_gc_blank_corr$OD600 - d_gc_blank_corr$blank_OD600
```

This correction gives occasional very small negative numbers, which is spurious - so I'll set those negative numbers to zero

```{r blank corr OD600 neg to zero}
d_gc_blank_corr_no_neg <- d_gc_blank_corr
d_gc_blank_corr_no_neg[(d_gc_blank_corr_no_neg$OD600_corr < 0), "OD600_corr"] <- 0
```


#Remove edge effect-y columns

```{r remove columns 2 and 11 since they are crashy}
d_gc_blank_corr_no_neg <- d_gc_blank_corr_no_neg %>%
  separate(col = well, 
           into = c("row", "column"),
           sep = 1,
           remove = FALSE) %>%
  filter(column != "2") %>%
  filter(column != "11")
```


#Plot growth curves

```{r generate summary data for plotting gc}
d_gc_corr_summary <- summarySE(data = d_gc_blank_corr_no_neg, measurevar = "OD600_corr", groupvars = c("time", "Condition"))
```


```{r plot growth curve mean ± SD}
p <- ggplot(data = d_gc_corr_summary, aes(x = time, y = OD600_corr, ymin = OD600_corr - sd, ymax = OD600_corr + sd, group = factor(Condition), color = factor(Condition), fill = factor(Condition), shape = factor(Condition)))

p <- p + geom_point(size = 2) + geom_errorbar() + 
  scale_color_manual(
    name = "Experimental Condition", 
    breaks = c("H", "G", "F", "E", "D", "C", "B", "A"),
    labels = c("2.7 mM C", "2.7 mM C (33% metabolite)", "1.8 mM C", "1.8 mM C (50% metabolite)", "0.90 mM C", "0.90 mM C (100% metabolite)", "0.45 mM C (100% metabolite)", "0 mM C"),
    values = c("#084594", "#4a1486", "#4292c6", "#807dba", "#9ecae1", "firebrick4", "#d73027", "gray70")) +
  scale_fill_manual(
    name = "Experimental Condition", 
    breaks = c("H", "G", "F", "E", "D", "C", "B", "A"),
    labels = c("2.7 mM C", "2.7 mM C (33% metabolite)", "1.8 mM C", "1.8 mM C (50% metabolite)", "0.90 mM C", "0.90 mM C (100% metabolite)", "0.45 mM C (100% metabolite)", "0 mM C"),
    values = c("#084594", "#4a1486", "#4292c6", "#807dba", "#9ecae1", "firebrick4", "#d73027", "gray70")) +
  scale_shape_manual(
    name = "Experimental Condition", 
    breaks = c("H", "G", "F", "E", "D", "C", "B", "A"),
    labels = c("2.7 mM C", "2.7 mM C (33% metabolite)", "1.8 mM C", "1.8 mM C (50% metabolite)", "0.90 mM C", "0.90 mM C (100% metabolite)", "0.45 mM C (100% metabolite)", "0 mM C"),
    values = c(24, 2, 23, 5, 22, 0, 1, 4)) + 
  theme_classic() + 
  xlab("Time (hrs)") + 
  ylab(bquote('OD'[600])) + 
  ggtitle(mtab_full_name) +
  theme(text = element_text(size = 16))

save_plots(p, "3m2op_corrected_gc.pdf", path_for_saving_plots)

p <- p + theme(legend.position =  "none")

save_plots(p, "3m2op_corrected_gc_noLegend.pdf", path_for_saving_plots)
```


#Iterate through wells and find the overall carrying capacity (max - min OD600)

```{r calculate carrying capacity}
#Set myself up for storing well calculations
well_index <- unique(d_gc_blank_corr_no_neg[c("well")])

d_gc_blank_corr_no_neg$minOD <- as.numeric(NA)
d_gc_blank_corr_no_neg$maxOD <- as.numeric(NA)

d_gc_capacityParams <- d_gc_blank_corr_no_neg %>%
  filter(well == "A")

#Find min and max OD600 in each well
for(i in well_index$well){
  d_temp <- d_gc_blank_corr_no_neg %>%
    filter(well == i)
  
  #find and add in min and max OD600
  minOD <- min(d_temp$OD600_corr)
  maxOD <- max(d_temp$OD600_corr)
  
  d_temp$minOD <- minOD
  d_temp$maxOD <- maxOD
  
  #bind values in
  d_gc_capacityParams <- rbind(d_gc_capacityParams, d_temp)
}

#Calculate carrying capacity
d_gc_capacityParams$K_overall <- d_gc_capacityParams$maxOD - d_gc_capacityParams$minOD

```

#Plot and test significance of condition on carrying capacity

```{r filter to one row per well}
#capacity parameters aren't changing with time, only need to represent those parameters once per well
d_capacity <- d_gc_capacityParams %>%
  filter(time == 0)
```


```{r test significance of condition on K_overall}
s_anova_K_overall <- aov(data = d_capacity, K_overall ~ Condition.name)
summary(s_anova_K_overall)
```

```{r post-hoc test}
TukeyHSD(s_anova_K_overall)
```

Condition does have a significant effect on carrying capacity. There is no significant difference between conditions with 3m2op and conditions with the equivalent amount of pyruvate but no 3m2op (i.e., 0 µM pyr ± 3m2op, N.S.; 300 µM pyr ± 3m2op, N.S.; 600 µM pyr ± 3m2op, N.S)

```{r plot K by condition}
p <- ggplot(data = d_capacity, aes(x = Condition.name, y = K_overall, color = Carbon_source))

p <- p + 
  geom_boxplot() + 
  geom_point() + 
   scale_color_manual(
     name = "Carbon source",
  #   breaks = c("3m2ob", "Pyruvate", "Pyruvate and 3m2ob", "None"),
  #   labels = c(mtab_full_name, "Pyruvate", "Mix", "No carbon added"),
     values = c("firebrick", "#084594", "#4a1486", "gray")) + 
  theme_classic(base_size = 16)

save_plots(p, "3m2op_K_byCondition_boxplot_and_points.pdf", path_for_saving_plots)

p <- p + theme(legend.position = "none")

save_plots(p, "3m2op_K_byCondition_boxplot_and_points_noLegend.pdf", path_for_saving_plots)
```

