---
title: "Alteromonas_growth_3m2ob"
author: "KHH"
date: "4/8/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals and overview

Analyze growth of Alteromonas macleodii on 3-methyl-2-oxobutanoic acid ± pyruvate. 

In this analysis we tested different ways of dealing with edge effects, which impacted the growth of Alteromonas grown on the outside edges of 96-well plates. We found that simply removing wells from the two rows/columns around the outside of the plate resulted in an insufficient sample size. Instead, we blank corrected, plotted wells individually within a culture condition, and removed edge rows/columns with wells which showed evidence of edge effects. Edge effects typically appeared as cultures which "crashed" immediately after reaching their max OD600, rather than entering stationary phase. 

We next blank corrected, plotted data, found the carrying capacity for each culture, and statistically compared carrying capacities between culture conditions. 


#Get likely required packages and functions


```{r get packages}
#for data management and plotting
require(tidyverse)

#some useful color palettes 
require(ggsci)
require(pals)
require(PaletteWoodsHole)
```


```{r get summary function}
## Summarizes data.
## Gives count, mean, standard deviation, standard error of the mean, and confidence interval (default 95%).
##   data: a data frame.
##   measurevar: the name of a column that contains the variable to be summariezed
##   groupvars: a vector containing names of columns that contain grouping variables
##   na.rm: a boolean that indicates whether to ignore NA's
##   conf.interval: the percent range of the confidence interval (default is 95%)
summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
    library(plyr)

    # New version of length which can handle NA's: if na.rm==T, don't count them
    length2 <- function (x, na.rm=FALSE) {
        if (na.rm) sum(!is.na(x))
        else       length(x)
    }

    # This does the summary. For each group's data frame, return a vector with
    # N, mean, and sd
    datac <- ddply(data, groupvars, .drop=.drop,
      .fun = function(xx, col) {
        c(N    = length2(xx[[col]], na.rm=na.rm),
          mean = mean   (xx[[col]], na.rm=na.rm),
          sd   = sd     (xx[[col]], na.rm=na.rm)
        )
      },
      measurevar
    )

    # Rename the "mean" column    
    datac <- rename(datac, c("mean" = measurevar))

    datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean

    # Confidence interval multiplier for standard error
    # Calculate t-statistic for confidence interval: 
    # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
    ciMult <- qt(conf.interval/2 + .5, datac$N-1)
    datac$ci <- datac$se * ciMult

    return(datac)
}

#example use:
#d_summary_name <- summarySE(d_example, measurevar = "variable to summarize", groupvars = c("group by this", "group by that"))
```


```{r set path and mtab names}
##Set your path for saving plots
path_for_saving_plots <- "~/Documents/MIT_WHOI_JP/Kuj:Chisholm data/Growth curves/04.2024_reanalysis/3m2ob/Figures/"

  
##Set your metabolite name
mtab_full_name <- "3m2ob"
mtab_short_name <- "3m2ob"

```


```{r define a function to save plots}
save_plots <- function(p, filename, path){
  ggsave(filename = filename, path = path, plot = p, device = "pdf", dpi = "retina", bg = "transparent")
}
```


## Get data

This analysis requires two dataframes: the OD600 data pulled from the plate reader, and a spreadsheet that describes culture condition by well position. 

```{r get data into R and rearrange}
#Edit the paths below to get to 1. raw OD600 data and 2. a data table that contains condition descriptions. That condition description file should include the following column headers: well, Condition.number, Condition, Condition.name, Total_uM_C, Total_mM_C, Carbon_source, Diauxic. 

OD600_data <- "~/Documents/MIT_WHOI_JP/Kuj:Chisholm data/Growth curves/Metabolite addition/3m2ob/3m2ob addition 7/KHH01_041_3m2ob_7_forR.csv"

culture_condition_data <- "~/Documents/MIT_WHOI_JP/Kuj:Chisholm data/Growth curves/Metabolite addition/3m2ob/3m2ob addition 7/3m2ob_7_condition_descriptions.csv"

#Now get those data into R and into useful formats

d_OD600 <- read.table(OD600_data, header = TRUE, sep = ",", stringsAsFactors = FALSE)
d_culture_conditions <- read.table(culture_condition_data, header = TRUE, sep = ",", stringsAsFactors = FALSE)

d_OD600_long <- gather(d_OD600, key = "well", value = "OD600", "A1", "A2", "A3", "A4", "A5", "A6", "A7", "A8", "A9", "A10", "A11", "A12", "B1", "B2", "B3", "B4", "B5", "B6", "B7", "B8", "B9", "B10", "B11", "B12", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "C9", "C10", "C11", "C12", "D1", "D2", "D3", "D4", "D5", "D6", "D7", "D8", "D9", "D10", "D11", "D12", "E1", "E2", "E3", "E4", "E5", "E6", "E7", "E8", "E9", "E10", "E11", "E12", "F1", "F2", "F3", "F4", "F5", "F6", "F7", "F8", "F9", "F10", "F11", "F12", "G1", "G2", "G3", "G4", "G5", "G6", "G7", "G8", "G9", "G10", "G11", "G12", "H1", "H2", "H3", "H4", "H5", "H6", "H7", "H8", "H9", "H10", "H11", "H12")

d_gc <- full_join(d_OD600_long, d_culture_conditions, by = "well")
```

#Filter out low-C mtab condition
Our experimental design included a lower concentration metabolite addition, but we decided it did not contribute to our understanding of Alteromonas metabolite uptake, and so that condition was not considered in the manuscript. 

Also have a look at filtering out edge wells, but it isn't a tractable approach

```{r filter out low-C mtab condition}
d_gc <- d_gc %>%
   filter(Total_mM_C != 0.45)
```

#Try a generic approch to removing edge effected wells 
It didn't work, but leaving here for the sake of record keeping. 

```{r trying a generic approach to dealing w edge effects}
#This does work, but helped me learn that if I just take away two outer rows/columns, I won't have sufficient sample size across all conditions. So the only way to deal with these data is by removing wells that visibly crash. 

# d_gc <- d_gc %>%
#   filter(Total_mM_C != 0.45) %>%
#   separate(col = well, 
#            into = c("row", "column"),
#            sep = 1, 
#            remove = FALSE) %>%
#   filter(row != "A") %>%
#   filter(row != "B") %>%
#   filter(row != "G") %>%
#   filter(row != "H") %>%
#   filter(column != "1") %>%
#   filter(column != "2") %>%
#   filter(column != "11") %>%
#   filter(column != "12")
```

#Blank correct

```{r make dataframe for data from blanks}
d_blank_OD600 <- filter(d_gc, Condition == "A")
```

```{r calculate mean OD600 at each t in blank}
d_summary_blank_OD600 <- summarySE(d_blank_OD600, measurevar = "OD600", groupvars = c("time"))
```


```{r add average blank values back into main dataframe and calc corrected OD600}
d_summary_blank_OD600 <- dplyr::rename(d_summary_blank_OD600, "blank_OD600" = "OD600")
d_summary_blank_OD600 <- dplyr::select(d_summary_blank_OD600, "time", "blank_OD600")
d_gc_blank_corr <- full_join(d_gc, d_summary_blank_OD600, by = "time")
d_gc_blank_corr$OD600_corr <- d_gc_blank_corr$OD600 - d_gc_blank_corr$blank_OD600
```

This correction gives occasional very small negative numbers, which is spurious - so set those negative numbers to zero

```{r blank corr OD600 neg to zero}
d_gc_blank_corr_no_neg <- d_gc_blank_corr
d_gc_blank_corr_no_neg[(d_gc_blank_corr_no_neg$OD600_corr < 0), "OD600_corr"] <- 0
#This worked, sick
```

#Plot individual cells to ID and remove edge effect crashers

```{r}
p <- ggplot(d_gc_blank_corr_no_neg, aes(x = time, y = OD600_corr, color = Condition.name))

p + 
  geom_point(size = 1.5) + 
  scale_color_lancet() + 
  labs(x = "Time (hrs)", y = bquote('OD'[600]), title = mtab_full_name, legend = "Condition name") +
  theme_classic(base_size = 16) 

```
Figure out which wells are crashy - in 300 uM pyr, 300 uM pyr + 180 uM 3m2ob, 900 uM Pyr wells


```{r}
d_gc_300uMpyr <- d_gc_blank_corr_no_neg %>% filter(Condition == "D")

p <- ggplot(d_gc_300uMpyr, aes(x = time, y = OD600_corr, color = well))

p + 
  geom_point(size = 1.5) + 
  scale_color_lancet() + 
  labs(x = "Time (hrs)", y = bquote('OD'[600]), title = "3m2ob experiment, 300 µM pyruvate") +
  theme_classic(base_size = 16) 

```

```{r}
d_gc_300uMpyr180mtab <- d_gc_blank_corr_no_neg %>% filter(Condition == "E")

p <- ggplot(d_gc_300uMpyr180mtab, aes(x = time, y = OD600_corr, color = well))

p + 
  geom_point(size = 1.5) + 
  scale_color_lancet() + 
  labs(x = "Time (hrs)", y = bquote('OD'[600]), title = "300 µM pyruvate 180 µM 3m2ob") +
  theme_classic(base_size = 16) 

```



```{r}
d_gc_600uMpyr <- d_gc_blank_corr_no_neg %>% filter(Condition == "F")

p <- ggplot(d_gc_600uMpyr, aes(x = time, y = OD600_corr, color = well))

p + 
  geom_point(size = 1.5) + 
  scale_color_lancet() + 
  labs(x = "Time (hrs)", y = bquote('OD'[600]), title = "3m2ob experiment, 600 µM pyruvate") +
  theme_classic(base_size = 16) 

```


```{r}
d_gc_900uMpyr <- d_gc_blank_corr_no_neg %>% filter(Condition == "H")

p <- ggplot(d_gc_900uMpyr, aes(x = time, y = OD600_corr, color = well))

p + 
  geom_point(size = 1.5) + 
  scale_color_lancet() + 
  labs(x = "Time (hrs)", y = bquote('OD'[600]), title = "900 µM pyruvate") +
  theme_classic(base_size = 16) 

```

Wells to remove: D11, E11, C2, C11

Leave B and G in (A and H were already buffer) and just remove columns 2 and 11?. Then I'm consistent all the way through, but removing as little as possible. 


```{r remove columns 2 and 11 since they are crashy}
d_gc_blank_corr_no_neg <- d_gc_blank_corr_no_neg %>%
  separate(col = well, 
           into = c("row", "column"),
           sep = 1,
           remove = FALSE) %>%
  filter(column != "2") %>%
  filter(column != "11")
```


```{r plot all data individually one more time}
p <- ggplot(d_gc_blank_corr_no_neg, aes(x = time, y = OD600_corr, color = Condition.name))

p + 
  geom_point(size = 1.5) + 
  scale_color_lancet() + 
  labs(x = "Time (hrs)", y = bquote('OD'[600]), title = mtab_full_name, legend = "Condition name") +
  theme_classic(base_size = 16) 

```

```{r get rid of the data sets that are just one condition to clean things up}
rm(d_gc_300uMpyr)
rm(d_gc_300uMpyr180mtab)
rm(d_gc_600uMpyr)
rm(d_gc_900uMpyr)
```

#Plot growth curves

```{r generate summary data for plotting gc}
d_gc_corr_summary <- summarySE(data = d_gc_blank_corr_no_neg, measurevar = "OD600_corr", groupvars = c("time", "Condition"))
```


```{r plot growth curve mean ± SD}
p <- ggplot(data = d_gc_corr_summary, aes(x = time, y = OD600_corr, ymin = OD600_corr - sd, ymax = OD600_corr + sd, group = factor(Condition), color = factor(Condition), fill = factor(Condition), shape = factor(Condition)))

p <- p + geom_point(size = 2) + geom_errorbar() + 
  scale_color_manual(
    name = "Experimental Condition", 
    breaks = c("H", "G", "F", "E", "D", "C", "B", "A"),
    labels = c("2.7 mM C", "2.7 mM C (33% metabolite)", "1.8 mM C", "1.8 mM C (50% metabolite)", "0.90 mM C", "0.90 mM C (100% metabolite)", "0.45 mM C (100% metabolite)", "0 mM C"),
    values = c("#084594", "#4a1486", "#4292c6", "#807dba", "#9ecae1", "firebrick4", "#d73027", "gray70")) +
  scale_fill_manual(
    name = "Experimental Condition", 
    breaks = c("H", "G", "F", "E", "D", "C", "B", "A"),
    labels = c("2.7 mM C", "2.7 mM C (33% metabolite)", "1.8 mM C", "1.8 mM C (50% metabolite)", "0.90 mM C", "0.90 mM C (100% metabolite)", "0.45 mM C (100% metabolite)", "0 mM C"),
    values = c("#084594", "#4a1486", "#4292c6", "#807dba", "#9ecae1", "firebrick4", "#d73027", "gray70")) +
  scale_shape_manual(
    name = "Experimental Condition", 
    breaks = c("H", "G", "F", "E", "D", "C", "B", "A"),
    labels = c("2.7 mM C", "2.7 mM C (33% metabolite)", "1.8 mM C", "1.8 mM C (50% metabolite)", "0.90 mM C", "0.90 mM C (100% metabolite)", "0.45 mM C (100% metabolite)", "0 mM C"),
    values = c(24, 2, 23, 5, 22, 0, 1, 4)) + 
  theme_classic() + 
  xlab("Time (hrs)") + 
  ylab(bquote('OD'[600])) + 
  ggtitle(mtab_full_name) +
  theme(text = element_text(size = 16))

save_plots(p, "3m2ob_corrected_gc.pdf", path_for_saving_plots)

p <- p + theme(legend.position =  "none")

save_plots(p, "3m2ob_corrected_gc_noLegend.pdf", path_for_saving_plots)
```

#Iterate through wells and find the overall carrying capacity (max - min OD600)

```{r calculate carrying capacity}
#Set myself up for storing well calculations
well_index <- unique(d_gc_blank_corr_no_neg[c("well")])

d_gc_blank_corr_no_neg$minOD <- as.numeric(NA)
d_gc_blank_corr_no_neg$maxOD <- as.numeric(NA)

d_gc_capacityParams <- d_gc_blank_corr_no_neg %>%
  filter(well == "A")

#Find min and max OD600 in each well
for(i in well_index$well){
  d_temp <- d_gc_blank_corr_no_neg %>%
    filter(well == i)
  
  #find and add in min and max OD600
  minOD <- min(d_temp$OD600_corr)
  maxOD <- max(d_temp$OD600_corr)
  
  d_temp$minOD <- minOD
  d_temp$maxOD <- maxOD
  
  #bind values in
  d_gc_capacityParams <- rbind(d_gc_capacityParams, d_temp)
}

#Calculate carrying capacity
d_gc_capacityParams$K_overall <- d_gc_capacityParams$maxOD - d_gc_capacityParams$minOD

```

#Plot and test significance of condition on carrying capacity

```{r filter to one row per well}
#capacity parameters aren't changing with time, only need to represent those parameters once per well
d_capacity <- d_gc_capacityParams %>%
  filter(time == 0)
```


```{r test significance of condition on K_overall}
s_anova_K_overall <- aov(data = d_capacity, K_overall ~ Condition.name)
summary(s_anova_K_overall)
```

```{r post-hoc test}
TukeyHSD(s_anova_K_overall)
```

Every level is significantly different from the other except for 300 µM pyruvate ± 180 µM 3m2ob. 

```{r plot K by condition}
p <- ggplot(data = d_capacity, aes(x = Condition.name, y = K_overall, color = Carbon_source))

p <- p + 
  geom_boxplot() + 
  geom_point() + 
   scale_color_manual(
     name = "Carbon source",
  #   breaks = c("3m2ob", "Pyruvate", "Pyruvate and 3m2ob", "None"),
  #   labels = c(mtab_full_name, "Pyruvate", "Mix", "No carbon added"),
     values = c("firebrick", "#084594", "#4a1486", "gray")) + 
  theme_classic(base_size = 16)

save_plots(p, "3m2ob_K_byCondition_boxplot_and_points.pdf", path_for_saving_plots)

p <- p + theme(legend.position = "none")

save_plots(p, "3m2ob_K_byCondition_boxplot_and_points_noLegend.pdf", path_for_saving_plots)
```

